---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE
)
```

# stagedtrees 

[![Build Status](https://travis-ci.com/gherardovarando/stagedtrees.svg?branch=master)](https://travis-ci.com/gherardovarando/stagedtrees)

### Installation 
  
```{r, eval = FALSE}
#development version from github
# install.packages("devtools")
devtools::install_github("gherardovarando/stagedtrees")

#unstable development version from the dev branch
devtools::install_github("gherardovarando/stagedtrees", ref = "dev")
```

### Usage

```{r}
library("stagedtrees")
```

With the `stagedtrees` package it is possible to fit (stratified) staged 
event trees to data, use them to compute probabilities, make predictions, 
visualize and compare different models. 

#### Creating the model

A staged event tree object (`sevt` class) can be created with the function
`staged_ev_tree`. In general we create a staged event tree from data 
in a `data.frame` or `table` object. 

```{r example}
# Load the Trump tweets data
data("Trump")

# Create the independence model 
indep <- staged_ev_tree(Trump, lambda = 1)
indep

#Create the full (saturated) model
full <- staged_ev_tree(Trump, full = TRUE, lambda = 1) 
full
```

#### Model selection

Starting from the independence model of the full model it is  
possible to perform automatic model selection. 

##### Score methods

This methods perform optimization of the model for a given score 
using different types of heuristic methods. 

* **Hill-Climbing** `hc.sevt(object, score, max_iter, trace)`
```{r}
mod1 <- hc.sevt(indep)
mod1
```
* **Backward Hill-Climbing** `bhc.sevt(object, score, max_iter, trace)`
```{r}
mod2 <- bhc.sevt(full)
mod2
```
* **Backward Fast Hill-Climbing** 
`fbhc.sevt(object, score, max_iter, trace)`
```{r}
mod3 <- fbhc.sevt(full, score = function(x) -BIC(x))
mod3
```


##### Distance methods

* **Backward Joining** `bj.sevt(full, distance, thr, trace, ...)`
```{r}
mod4 <- bj.sevt(full)
mod4
```

* **Naive model** `naive.sevt(full, distance, k)`
```{r}
mod5 <- naive.sevt(full)
mod5
```
#### Probabilities, predictions and sampling

##### Marginal probabilities

Obtain marginal probabilities with the  `prob.sevt` function.
```{r}
# estimated probability of (Source = "iOS", Sentiment = "Negative")
# using different models
prob.sevt(indep, c(Source = "iOS", Sentiment = "Negative")) 
prob.sevt(mod3, c(Source = "iOS", Sentiment = "Negative"))
```

Or for a  `data.frame` of observations:
```{r}
obs <- expand.grid(full$tree[c(2,3,5)])
p <- prob.sevt(mod2, obs)
cbind(obs, P = p)
```

##### Predictions

A staged event tree object can be used to make predictions with the
`predict` method. The class variable can be specified, otherwise the 
first variable (root) in the tree will be used. 

```{r}
predicted <- predict(mod3, newdata = Trump)
table(predicted, Trump$Source)
```

##### Sampling

```{r}
sample.sevt(mod4, 5)
```
#### Explore the model

##### Plot 

```{r}
plot(mod4, main = "Staged tree learned with bj.sevt", 
     cex.label.edges = 0.6, cex.nodes = 1.5)
text(mod4, y = -0.03, cex = 0.7)
```

##### Stages

```{r}
stageinfo.sevt(mod4, var = "Time")
```

##### Subtrees

Subtrees can be extracted, the result is another staged event tree object
in the remaining variables.

```{r}
sub <- subtree.sevt(mod4, c("iOS", ">20"))
plot(sub)
text(sub, y = -0.03, cex = 0.7)
```

#### Comparing models

Check if the models are equal.
```{r}
compare.sevt(mod1, mod2)
```

Penalized log-likelihood.
```{r}
BIC(mod1, mod2, mod3, mod4, mod5)
```