---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# stagedtrees

[![stagedtrees](https://www.r-pkg.org/badges/version/stagedtrees)](https://cran.r-project.org/package=stagedtrees)
  [![R-CMD-check](https://github.com/stagedtrees/stagedtrees/workflows/R-CMD-check/badge.svg)](https://github.com/stagedtrees/stagedtrees/actions)
[![codecov](https://codecov.io/github/stagedtrees/stagedtrees/branch/main/graph/badge.svg?token=K76ZINC3IT)](https://codecov.io/github/stagedtrees/stagedtrees)
 [![](https://cranlogs.r-pkg.org/badges/stagedtrees)](https://cran.r-project.org/package=stagedtrees)

##### Citation

To cite stagedtrees in publications use:

Carli F, Leonelli M, Riccomagno E, Varando G (2022). “The R Package stagedtrees for
Structural Learning of Stratified Staged Trees.” _Journal of Statistical Software_,
*102*(6), 1-30. doi: 10.18637/jss.v102.i06 (URL:
https://doi.org/10.18637/jss.v102.i06).

```
@Article{,
    title = {The {R} Package {stagedtrees} for Structural Learning of Stratified Staged Trees},
    author = {Federico Carli and Manuele Leonelli and Eva Riccomagno and Gherardo Varando},
    journal = {Journal of Statistical Software},
    year = {2022},
    volume = {102},
    number = {6},
    pages = {1--30},
    doi = {10.18637/jss.v102.i06},
  }
```

### Overview 

`stagedtrees` is a package that implements staged event trees, a class of
probability models for categorical random variables.   

### Installation 
  
```{r, eval = FALSE}
# Install stable version from CRAN:
install.packages("stagedtrees")

# Or the development version from GitHub:
remotes::install_github("stagedtrees/stagedtrees")
```

### Usage

With the `stagedtrees` package it is possible to estimate (stratified) staged event trees from data, use them to compute probabilities, make predictions, visualize and compare different models. 

```{r}
library("stagedtrees")

tree <- Titanic |> full() |> stages_bhc() |> stndnaming(uniq = TRUE)

prob(tree, c(Survived="Yes"), conditional_on = c(Age="Adult"))

palette("Okabe-Ito")
par(mfrow = c(1,2))
plot(tree, col = "stages")
barplot(tree, var = "Survived", main = "P(Survived|-)", col = "stages")
```



