---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# stagedtrees

[![stagedtrees](https://www.r-pkg.org/badges/version/stagedtrees)](https://cran.r-project.org/package=stagedtrees)
[![Build Status](https://travis-ci.com/gherardovarando/stagedtrees.svg?branch=main)](https://app.travis-ci.com/github/gherardovarando/stagedtrees)
  [![R-CMD-check](https://github.com/gherardovarando/stagedtrees/workflows/R-CMD-check/badge.svg)](https://github.com/gherardovarando/stagedtrees/actions)
 [![Coverage status](https://codecov.io/gh/gherardovarando/stagedtrees/branch/master/graph/badge.svg)](https://codecov.io/github/gherardovarando/stagedtrees?branch=main)
 [![](https://cranlogs.r-pkg.org/badges/stagedtrees)](https://cran.r-project.org/package=stagedtrees)

##### Preprint

F Carli, M Leonelli, E Riccomagno, G Varando, The R Package stagedtrees for Structural Learning of Stratified Staged Trees, 2020
[arXiv:2004.06459](https://arxiv.org/abs/2004.06459)

```
@misc{2004.06459,
Author = {Federico Carli and Manuele Leonelli and Eva Riccomagno and Gherardo Varando},
Title = {The R Package stagedtrees for Structural Learning of Stratified Staged Trees},
Year = {2020},
Eprint = {arXiv:2004.06459},
}
```

### Overview 

`stagedtrees` is a package that implements staged event trees, a probability
model for categorical random variables.   

### Installation 
  
```{r, eval = FALSE}
#stable version from CRAN 
install.packages("stagedtrees")

#development version from github
# install.packages("devtools")
devtools::install_github("gherardovarando/stagedtrees")
```

### Usage

```{r}
library("stagedtrees")
```

With the `stagedtrees` package it is possible to fit (stratified) staged 
event trees to data, use them to compute probabilities, make predictions, 
visualize and compare different models. 


#### Creating the model

A staged event tree object (`sevt` class) 
can be initialized as the full (saturated) or as 
the fully independent model with, respectively, 
the functions `indep` and `full`. 
It is possible to build a staged event tree 
from data stored in a `data.frame` or a 
`table` object. 

```{r example}
# Load the PhDArticles data
data("Titanic")

# define order of variables
order <- c("Sex", "Age",  "Class", "Survived")

# Independence model 
mod_indep <- indep(Titanic, order)
mod_indep

# Full (saturated) model
mod_full <- full(Titanic, order) 
mod_full
```

##### Structural zeros and unobserved situations

By default staged trees object are defined assuming structural zeros 
in the contingency tables. This is implemented by joining all 
unobserved situations in particular stages (named by default `"UNOBSERVED"`) 
which are, by default, 
ignored by other methods and functions (see the `ignore` argument in 
`?stages_bhc` or `?plot.sevt`).

```{r}
## there are no observations for Sex=Male (Female), Age = Child, Class = Crew
get_stage(mod_full, c("Male", "Child", "Crew"))

## and obviously 
prob(mod_full, c(Age = "CHild", CLass = "Crew"))
```

###### Initialize a model without structural zeros 

It is possible to initialize a staged tree without structural zeros 
by setting the argument `join_unobserved=FALSE`. 
In that case, it can be useful to set `lambda > 0` to avoid problems 
with probabilities on unobserved situations. 

```{r}
mod_full0 <- full(Titanic, join_unobserved = FALSE, lambda = 1)
```

#### Model selection

`stagedtrees` implements methods to perform automatic model selection. All methods can be initialized from an arbitrary staged event tree object. 

##### Score methods

This methods perform optimization for a given score 
using different heuristics. 

* **Hill-Climbing** `stages_hc(object, score, max_iter, scope, ignore, trace)`
```{r}
mod1 <- stages_hc(mod_indep)
mod1
```
* **Backward Hill-Climbing** `stages_bhc(object, score, max_iter, scope, ignore, trace)`
```{r}
mod2 <- stages_bhc(mod_full)
mod2
```

* **Backward Fast Hill-Climbing** `stages_fbhc(object, score, max_iter, scope, ignore, trace)`
```{r}
mod3 <- stages_fbhc(mod_full, score = function(x) -BIC(x))
mod3
```

##### Clustering methods

* **Backward Joining** `stages_bj(object, distance, thr, scope, ignore, trace)`
```{r}
mod4 <- stages_bj(mod_full)
mod4
```

* **Hierarchical Clustering** `stages_hclust(object, distance, k, method, ignore, limit, scope)`
```{r}
mod5 <- stages_hclust(mod_full,
                    k = 2, 
                    distance = "totvar",
                   method = "mcquitty")
mod5
```

* **K-Means Clustering** `stages_kmeans(object, k, algorithm, ignore, limit, scope, nstart)`
```{r}
mod6 <- stages_kmeans(mod_full,
                    k = 2, 
                   algorithm = "Hartigan-Wong")
mod6
```

#### Combining model selections with `|>` (or `%>%`)

The new native pipe operator `|>` (or the one  from the `magrittr` package) can be used to combine
various model selection algorithms. 

```{r}
model <- Titanic |> full(lambda = 1) |> stages_hclust() |> stages_hc()

## extract a sub_tree and join two stages
small_model <- model |> subtree(path = c("Crew"))  |>
              join_stages("Survived", "3", "7")
```

#### Probabilities, predictions and sampling

##### Marginal probabilities

Obtain marginal (or conditionals) probabilities with the  `prob` function.
```{r}
# estimated probability of c(Sex = "Male", Class = "1st")
# using different models
prob(mod_indep, c(Sex = "Male", Class = "1st")) 
prob(mod3, c(Sex = "Male", Class = "1st"))
```

Or for a  `data.frame` of observations:

```{r}
obs <- expand.grid(mod_full$tree[c(1,3)])
p <- prob(mod2, obs)
cbind(obs, P = p)
```

Conditional probabilities can be obtained via the 
`conditional_on` argument. 
```{r}
prob(mod3, c(Sex = "Male", Class = "1st"),       
     conditional_on = c(Survived = "Yes"))
```

##### Predictions

A staged event tree object can be used to make predictions with the
`predict` method. The class variable can be specified, otherwise the 
first variable (root) in the tree will be used. 

```{r}
## check accuracy over the Titanic data
titanic_df <- as.data.frame(Titanic)
predicted <- predict(mod3, class = "Survived", newdata = titanic_df)
table(predicted, titanic_df$Survived)
```

Conditional probabilities (or log-) can be obtained setting `prob = TRUE`:

```{r}
## obtain estimated conditional probabilities in mod3 
predict(mod3, newdata = titanic_df[1:3,], prob = TRUE)
```

##### Sampling

```{r}
sample_from(mod4, 5)
```

#### Explore the model

##### Model info

```{r}
# stages
stages(mod1, "Age")

# summary
summary(mod1)

# confidence intervals
confint(mod1, parm = "Age")
```

##### Plot 

```{r}
plot(mod4, main = "Staged tree learned with bj.sevt", 
     cex_label_edges = 0.6, cex_nodes = 1.5)
```

By default stages associated with the unobserved situations are not plotted, 
if the model has been created with `join_unobserved = TRUE`. 
But we can plot also the unobserved stages in a specific color.
```{r}
plot(stndnaming(mod5, uniq = TRUE), 
     main = "Staged tree learned with stages_hclust 
     (unobserved in grey)",  
     ignore = FALSE, ## do not ignore stages
     col = function(stages) ifelse(stages=="UNOBSERVED", "grey", stages),
     cex_label_edges = 0.6, cex_nodes = 1.5)
```

###### Barplot

The method `barplot.sevt` creates a bar plot 
for the conditional probabilities of a variable.

```{r}
barplot(mod4, "Class", legend.text = TRUE)
```

###### CEG plots

Plotting CEG requires the `igraph` package. 

```{r}
plot(ceg(mod5))
```

##### Subtrees

From a staged evnt tree a subtree can be extracted, 
the resulting model is ar staged event tree
in the remaining variables.

```{r}
sub <- subtree(mod4, c("Female"))
plot(sub)
```

#### Comparing models

##### Compare stages structure
```{r}
compare_stages(mod1, mod4)

compare_stages(mod1, mod4, method = "hamming", plot = TRUE, 
             cex_label_nodes = 0, cex_label_edges = 0)

hamming_stages(mod1, mod4)

difftree <- compare_stages(mod1, mod4, method = "stages", plot = FALSE, 
             return_tree = TRUE)

difftree$Married
```

##### Penalized log-likelihood.
```{r}
BIC(mod_indep, mod_full, mod1, mod2, mod3, mod4, mod5)
```

##### Likelihood-ratio test

```{r}
mod1a <- join_stages(mod1, "Class", "3", "4")
lr_test(mod1a, mod1)
```