---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# stagedtrees

[![Build Status](https://travis-ci.com/gherardovarando/stagedtrees.svg?branch=master)](https://travis-ci.com/gherardovarando/stagedtrees)
 [![Coverage status](https://codecov.io/gh/gherardovarando/stagedtrees/branch/master/graph/badge.svg)](https://codecov.io/github/gherardovarando/stagedtrees?branch=master)
 
 
### Overview 

`stagedtrees` is a package that implements staged event trees, a probability
model for discrete random variables.   

### Installation 
  
```{r, eval = FALSE}
#development version from github
# install.packages("devtools")
devtools::install_github("gherardovarando/stagedtrees")
```

### Usage

```{r}
library("stagedtrees")
```

With the `stagedtrees` package it is possible to fit (stratified) staged 
event trees to data, use them to compute probabilities, make predictions, 
visualize and compare different models. 


#### Creating the model

A staged event tree object (`sevt` class) can be created with the function
`staged_ev_tree`, or with the functions `indep` and `full`. In general we create a staged event tree from data 
in a `data.frame` or `table` object. 

```{r example}
# Load the PhDArticles data
data("PhDArticles")

# Create the independence model 
mod_indep <- indep(PhDArticles, lambda = 1)
mod_indep

#Create the full (saturated) model
mod_full <- full(PhDArticles, lambda = 1) 
mod_full
```

#### Model selection

Starting from the independence model of the full model it is  
possible to perform automatic model selection. 

##### Score methods

This methods perform optimization of the model for a given score 
using different types of heuristic methods. 

* **Hill-Climbing** `hc.sevt(object, score, max_iter, trace)`
```{r}
mod1 <- hc.sevt(mod_indep)
mod1
```
* **Backward Hill-Climbing** `bhc.sevt(object, score, max_iter, trace)`
```{r}
mod2 <- bhc.sevt(mod_full)
mod2
```
* **Backward Fast Hill-Climbing** 
`fbhc.sevt(object, score, max_iter, trace)`
```{r}
mod3 <- fbhc.sevt(mod_full, score = function(x) -BIC(x))
mod3
```


##### Distance methods

* **Backward Joining** `bj.sevt(full, distance, thr, trace, ...)`
```{r}
mod4 <- bj.sevt(mod_full)
mod4
```

* **Naive model** `naive.sevt(full, distance, k)`
```{r}
mod5 <- naive.sevt(mod_full)
mod5
```

#### Combining model selections with `%>%`

The pipe operator from the `magrittr` package can be used to combine
easily various model selection algorithms and to specify models easily. 

```{r}
library(magrittr)
model <- PhDArticles %>% full(lambda = 1) %>% naive.sevt %>% 
               hc.sevt

## extract a sub_tree and join two stages
sub_model <- model %>% subtree.sevt(path = c(">2"))  %>%  
              join_stages("Mentor", "1", "2")
```


#### Probabilities, predictions and sampling

##### Marginal probabilities

Obtain marginal probabilities with the  `prob.sevt` function.
```{r}
# estimated probability of c(Gender = "male", Married = "yes")
# using different models
prob.sevt(mod_indep, c(Gender = "male", Married = "yes")) 
prob.sevt(mod3, c(Gender = "male", Married = "yes"))
```

Or for a  `data.frame` of observations:
```{r}
obs <- expand.grid(mod_full$tree[c(2,3,5)])
p <- prob.sevt(mod2, obs)
cbind(obs, P = p)
```

##### Predictions

A staged event tree object can be used to make predictions with the
`predict` method. The class variable can be specified, otherwise the 
first variable (root) in the tree will be used. 

```{r}
## check accuracy over the PhDArticles data
predicted <- predict(mod3, newdata = PhDArticles)
table(predicted, PhDArticles$Articles)
```

Conditional probabilities (or log-) can be obtained setting `prob = TRUE`:

```{r}
## obtain estimated conditional probabilities in mod3 for first 5 obs
## P(Articles|Gender, Kids, Married, Mentor, Prestige)
predict(mod3, newdata = PhDArticles[1:5,], prob = TRUE)
```
##### Sampling

```{r}
sample.sevt(mod4, 5)
```
#### Explore the model

##### Model info

```{r}
# Degrees of freedom
df.sevt(mod_full)
df.sevt(mod_indep)

# variables 
varnames.sevt(mod1)

# number of variables
nvar.sevt(mod1)
```

##### Plot 

```{r}
plot(mod4, main = "Staged tree learned with bj.sevt", 
     cex.label.edges = 0.6, cex.nodes = 1.5)
text(mod4, y = -0.03, cex = 0.7)
```

##### Stages

```{r}
stages.sevt(mod4, "Kids")
```

```{r}
summary(mod4)
```

##### Subtrees

A subtree can be extracted, the result is another staged event tree object
in the remaining variables.

```{r}
sub <- subtree.sevt(mod4, c(">2", "female"))
plot(sub)
text(sub, y = -0.03, cex = 0.7)
```

#### Comparing models

Check if models are equal.
```{r}
compare.sevt(mod1, mod2)

compare.sevt(mod1, mod2, method = "hamming", plot = TRUE, 
             cex.label.nodes = 0, cex.label.edges = 0)
text(mod1)

hamming.sevt(mod1, mod2)

difftree <- compare.sevt(mod1, mod2, method = "stages", plot = FALSE, 
             return.tree = TRUE)

difftree$Married
```

Penalized log-likelihood.
```{r}
BIC(mod_indep, mod_full, mod1, mod2, mod3, mod4, mod5)
```