---
title: "Agent-Assisted Defect Detection and Optimisation in an R Package: A Case Study on stagedtrees"
author:
  - name: Gherardo Varando
    affiliation: Universitat de València
    email: gherardo.varando@uv.es
abstract: >
  Large language model agents are increasingly proposed for software
  maintenance, but published evidence is dominated by synthetic benchmarks
  rather than sustained work on real, released software. We report a case
  study in which an LLM agent (Claude, Anthropic) reviewed and optimised
  **stagedtrees**, a CRAN package of roughly 6,800 lines implementing staged
  event tree models. Over a single extended session the agent identified 14
  latent defects, all reproduced before being reported, and applied a series
  of performance changes that leave the package's central fitting routine
  20.8 times faster and its structure-search algorithm 5.9 times faster, with
  output verified identical. A version-to-version benchmark also shows the
  correctness fixes carrying a measurable cost of their own — two query
  functions became 8 to 14 per cent slower before any optimisation was
  attempted — a trade-off that the usual framing of such work obscures.
  We give equal weight to the agent's failures:
  it introduced a regression caught only by the package's existing test
  suite, proposed an "obvious" optimisation that would have silently changed
  a stored attribute's type, omitted a namespace import, and — most
  instructively — made confident but incorrect claims about which functions
  mattered, which no automated check could have caught and which were
  corrected by the maintainer's domain knowledge. We argue that the
  distinction between these two failure classes is the central practical
  finding: mechanical verification reliably catches behavioural regressions,
  but judgements about *relevance* remain irreducibly human.
output:
  rticles::rjournal_article:
    keep_tex: true
bibliography: references.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
options(stringsAsFactors = FALSE)

bench_dir <- Sys.getenv("BENCH_DIR", unset = "../benchmarks/results")
have_bench <- dir.exists(bench_dir) &&
  length(list.files(bench_dir, pattern = "^bench-v[0-9]+\\.csv$")) >= 2

if (have_bench) {
  files <- list.files(bench_dir, pattern = "^bench-v[0-9]+\\.csv$", full.names = TRUE)
  bench <- do.call(rbind, lapply(files, read.csv))
  agg <- aggregate(seconds ~ version + workload + size, bench, median)
  wide <- reshape(agg, idvar = c("workload", "size"), timevar = "version",
                  direction = "wide")
  names(wide) <- sub("^seconds\\.", "", names(wide))
  if (all(c("v0", "v2") %in% names(wide))) {
    wide$speedup <- wide$v0 / wide$v2
    wide <- wide[order(-wide$speedup), ]
  }
}
```

# Introduction

Automated program repair and LLM-based coding assistants are an active
research area, but the evidence base is skewed. Benchmarks such as SWE-bench
measure an agent's ability to reproduce a known patch for a known issue;
they say little about the open-ended task a package maintainer actually
faces, which is to find defects nobody has reported yet and to make code
faster without changing what it computes.

This paper reports a case study of that open-ended task. The subject is
**stagedtrees** [@carli2022], a CRAN package implementing staged event tree
models — asymmetric graphical models for categorical data. The package is
roughly 6,800 lines of R across 65 files, has been on CRAN since 2020, and
is actively maintained.

We make three contributions. First, we characterise the defects an agent
found in released, tested code, and the discipline that made those findings
trustworthy. Second, we report performance work with verified-identical
output and quantify it across package versions. Third — and we regard this
as the most transferable result — we document the agent's failures in
detail and separate them into two classes with very different practical
consequences.

A necessary caveat: this paper describes work performed by an LLM agent,
and the first draft of this manuscript was produced by that same agent.
Self-report is a weak form of evidence. We have tried to mitigate this by
making every quantitative claim reproducible from scripts in the package
repository, and by giving failures at least as much space as successes.
Section \ref{sec:limitations} returns to this.

# Setting and method

## The package

stagedtrees fits staged event trees: models in which the vertices of an
event tree are partitioned into *stages* that share a conditional
probability distribution. The computational core is a greedy search over
stage partitions, scored by a penalised likelihood.

## Protocol

Work proceeded in three phases over one extended session:

1. **Review.** Four parallel sub-agents examined disjoint parts of the
   codebase. The governing instruction was: *do not report a finding you
   have not reproduced.* Every candidate defect had to be demonstrated with
   a concrete failing input before it could be reported.
2. **Triage.** The maintainer approved findings individually. Each approved
   fix became a separate commit with a regression test, so that any single
   fix could be reverted without the others.
3. **Optimisation.** Profiling directed the work; every change was required
   to demonstrate identical output, not merely similar output.

The third requirement did most of the load-bearing work, and is discussed
in Section \ref{sec:failures}.

# Defects found

Fourteen defects were reproduced and fixed. They fall into recognisable
classes rather than being idiosyncratic.

## Sequence-construction errors

Seven defects shared one root cause: `1:n` where `seq_len(n)` was meant.
When `n` is zero, `1:0` yields `c(1, 0)` and the loop runs twice with an
out-of-range index, rather than not running at all. In stagedtrees this
produced a crash in `sample_from()` on a single-variable model, a crash in
`prob()` on zero-row input, and — most seriously — *silent* double
iteration in two expectation-maximisation loops, where `for (i in
1:max_iter)` with `max_iter = 0` performed two EM iterations and changed
the fitted probabilities rather than leaving them untouched.

This class is worth highlighting because it is invisible to ordinary
testing: the failure requires a boundary input (an empty vector, a
single-variable model, a zero iteration count) that a test suite organised
around typical usage will never generate.

## Silent numerical corruption

Three defects produced wrong numbers rather than errors.

`prob()` returned `1` for every conditional query on a zero-probability
conditioning event. The mechanism: the probability was computed in log
space, where `log(0) - log(0)` gives `NaN`; a guard intended to map missing
values to zero mapped that `NaN` to `0`, and `exp(0)` is `1`. A query whose
correct answer is undefined returned the maximum possible probability.

`rename_stage()` silently merged two stages when the target name already
existed, overwriting one stage's probability vector with another's and
leaving the cached log-likelihood stale.

`indep.data.frame()` dropped missing values per-variable while the
contingency-table constructor used complete cases, so the same model fitted
through two code paths reported different log-likelihoods.

## Structural assumptions

The largest single fix concerned an undocumented convention. The `$stages`
component is a named list with one entry per non-root variable, but a root
entry is *optional* — some constructors add it, others do not. Five
functions indexed `$stages` positionally, and therefore computed silently
wrong results whenever the optional entry was present. Degrees of freedom,
and hence every BIC-based model comparison, were affected.

# Performance

```{r perf-table, results = "asis"}
if (have_bench && "speedup" %in% names(wide)) {
  tab <- wide[, c("workload", "size", "v0", "v1", "v2")]
  tab$fix_cost <- sprintf("%+.1f%%", 100 * (wide$v1 / wide$v0 - 1))
  tab$speedup <- sprintf("%.2f", wide$speedup)
  names(tab) <- c("Workload", "Size", "v0 (s)", "v1 (s)", "v2 (s)",
                  "v1 vs v0", "v0/v2")
  print(knitr::kable(tab, digits = 3, row.names = FALSE, booktabs = TRUE,
    caption = paste("Median elapsed time over repeated runs.",
      "v0 is the package before this work, v1 after the defect fixes,",
      "v2 after the optimisations.")))
} else {
  cat("_Benchmark results not found; run `benchmarks/run-versions.sh` first._\n")
}
```

Two features of Table 1 deserve comment before the individual changes.

First, the two smallest workloads (`full` at 5000 observations, and
`stages_hclust`) run in well under a tenth of a second and their ratios are
dominated by measurement noise; they are retained to show that nothing
regressed, not as evidence of improvement.

Second, and more interesting: **the defect fixes were not free**. Comparing
v1 against v0 — a comparison in which no performance work had yet been
done — `predict()` became 13.7% slower and `prob()` 7.5% slower, both well
outside the run-to-run spread. These are precisely the two functions most
affected by the correctness fixes described in Section 3: the zero-probability
guard added a `NaN` detection pass over the result vector, and the change
from positional to name-based indexing of `$stages` replaced integer
subscripting with string lookup. The correctness of the result improved and
the cost of obtaining it rose. We report this because the framing of AI
agents as delivering simultaneous correctness and speed improvements
obscures that these can trade against one another; here the subsequent
optimisation work more than recovered the loss, but it need not have.

Profiling, rather than intuition, directed every optimisation. Three
findings dominated.

**Fitting grouped situations quadratically.** `sevt_fit()` assigned
situations to stages by scanning the entire stage vector once per stage.
This is $O(\text{stages} \times \text{situations})$, and a saturated model
has one stage per situation, so the deepest variable of an eight-variable,
four-level model performed $16384^2 \approx 2.7 \times 10^8$ string
comparisons. Partitioning in a single pass with `split()` removed the
quadratic term. Because `full()` is the entry point for essentially every
workflow in the package, this was the most valuable change made.

**The search evaluated candidates by materialising them.** The backward
hill-climbing search scored each candidate stage merge by constructing a
complete modified model object — copying the stage vector, rebuilding the
nested probability list, carrying the contingency tables — and then reading
two scalars out of it, discarding the object for every candidate but the
winner. Since the score is a function of the log-likelihood and the degrees
of freedom alone, and both change by an amount computable in closed form
from the two probability vectors being merged, candidates can be scored
without building anything.

This required an interface decision. The `score` argument is user-supplied
and receives a whole model, so the fast path cannot be applied
unconditionally. The resolution was to let `score` be *either* a string
naming a predefined score (`"BIC"`, the new default, or `"AIC"`) or a
function; strings take the incremental path, functions retain the original
behaviour exactly. Users who supply custom scores are unaffected.

**Index computation recomputed model-invariant quantities.** The function
mapping a root-to-leaf path to an integer index recomputed the vector of
variable arities on every call — 17 of the 65 microseconds a call took —
for a quantity fixed by the model, and rebuilt a stride vector that is a
cumulative product using a quadratic pattern.

## What Rcpp was, and was not, worth

Because the package contains no compiled code, we measured whether it
should. A no-op compiled function called from R costs approximately 3
microseconds on the test machine. This exceeds the work done by any of the
individual helper functions identified above: a C++ implementation of the
path-index function measured 2.70 microseconds against 12 microseconds for
an optimised R implementation — a fourfold gain that would purchase a
`src/` directory, a `LinkingTo` dependency and a permanent cross-platform
build burden.

The calculus inverts when an entire loop moves into C rather than a single
function. Generating 20,000 sampled paths over 8 variables in one call
took 0.007 seconds against 16.0 seconds for the R implementation, because
the per-call overhead is paid once instead of 160,000 times. The
actionable rule is *port loops, not leaves*, and on that basis we
implemented none of it: the pure-R changes captured enough of the
available gain that compiled code was not justified.

# Where the agent was wrong {#sec:failures}

The agent's errors divide into two classes, and the distinction between
them is the practical finding of this study.

## Class 1: behavioural regressions, caught mechanically

Four errors changed program behaviour and were all caught by automated
means before reaching the repository.

**An omitted namespace import.** A fix introduced `complete.cases()` and
`setNames()` without declaring them, which `R CMD check` reported as
undefined globals. Caught by the standard package check.

**A reintroduced boundary bug.** The rewritten path-index function indexed
`path[[k]]` with `k = 0` for an empty path, where the previous
implementation returned `NA`. This is notable: the agent had verified the
rewrite exhaustively over all 396 paths of four trees, and the check missed
the case because the enumeration used `expand.grid()`, which never produces
a zero-length path. The regression was caught by a test the *same agent*
had added earlier in the session, for a defect in the same class. An
exhaustive check is only as good as its enumeration.

**A plausible optimisation that was neither.** Replacing `apply(x, 2, sum)`
with `colSums(x)` is a standard R idiom. Here it gained nothing — measured
at 1.0 times, because the cost was the surrounding scan, not the
summation — and `colSums()` returns double where `apply()` returns integer,
which would have silently changed the storage type of a sample-size
attribute on every fitted probability vector. Caught by comparing 192
fitted models with `identical()` rather than `all.equal()`.

**A dropped incidental attribute.** A vectorised replacement for a
row-filling loop lost names that the original `array()` call had
incidentally attached to a `dim` attribute. Semantically irrelevant;
caught by the same strict comparison.

The common thread is that strict structural comparison over many inputs
caught all four, whereas value comparison (`all.equal`) would have caught
only the first two. The cost of that strictness was low and its yield was
high.

## Class 2: misjudged relevance, caught only by the maintainer

The second class is more troubling, because no test could have caught it.

The agent profiled `stages_bj()` at 21.6 seconds and identified a genuine
superquadratic scaling defect in it — a distance matrix rebuilt in full
after every merge when only one row changes. It recommended the fix. It
also asserted that related changes would accelerate `stages_hclust()`.

The maintainer's response was that `stages_hclust()` uses R's built-in
`dist()` for its documented distance options and never calls the affected
code at all, and that the paths in question are not ones users exercise.
Both points were correct; the agent's claim about `stages_hclust()` was
simply false, and verifiable in seconds by grep — which the agent performed
only after being challenged, confirming the maintainer. The recommended
work was abandoned.

The agent had made a similar error earlier, reporting that four search
functions shared a code pattern when one of them — which takes no score
argument at all — did not.

These were not failures of reasoning about code behaviour, which the agent
did reliably. They were failures of judgement about *what matters*: which
functions are used, which costs users actually pay, where optimisation
effort earns its risk. That knowledge lives in the maintainer's experience
of the user base and is not recoverable from the source tree.

The redirection was consequential. Abandoning `stages_bj()` and
investigating `full()` instead led directly to the largest result in this
paper.

# Discussion

The two failure classes suggest an asymmetric division of labour.

For behavioural correctness, the agent was effective and mechanical
verification was sufficient. The productive discipline was not
sophisticated: reproduce before reporting, compare with `identical()`
rather than `all.equal()`, compare across many randomised inputs rather
than one, and treat any diagnosed difference as a reason to stop rather
than a detail to explain away. Every Class 1 error was caught this way,
and the cost was a few minutes of compute per change.

For relevance, the agent was actively misleading, because its errors were
delivered with the same confidence as its correct findings. An agent that
profiles a function at 21.6 seconds and proposes a correct algorithmic fix
gives no signal that the function is rarely called. Nothing in the
repository encodes that. A maintainer reading the same report identifies it
immediately.

This suggests that the review burden on a human collaborator does not scale
down in proportion to the agent's competence. The agent can be trusted to
establish *that* a change is safe; it cannot be trusted to establish that
the change is *worth making*, and the second question is where the
maintainer's attention is best spent.

# Limitations {#sec:limitations}

This is a single case study, and its limitations are substantial.

**Self-report.** The work was performed by an LLM agent and the first draft
of this paper was written by that agent. An agent summarising its own
performance is a poor instrument, particularly for detecting errors it did
not notice making. The failures documented in Section \ref{sec:failures}
are those that were *caught*; failures that passed both automated checks
and maintainer review would by construction be absent from this account.

**No control.** We do not know what a human reviewer, a static analyser, or
a different agent would have found in the same code. The defect count is
therefore not a measure of anything comparative. We note that several
defects — particularly the `1:n` family — are exactly what existing lint
tooling targets, and would plausibly have been found by cheaper means.

**Benchmark conditions.** Timings were collected in a single containerised
environment on one machine. Absolute figures should not be compared across
systems; the ratios are the meaningful quantity, and even those carry
run-to-run variance that the repeated-measures design exposes but does not
eliminate.

**Defect severity is unmeasured.** The defects found were latent: none
corresponded to a user-reported failure. Their practical importance is
therefore unknown, and counting them conflates a silent EM-iteration bug
with a redundant loop.

**Generalisation.** One package, one domain, one agent, one session. The
codebase is mid-sized, pure R, and has an unusually well-defined numerical
contract, which makes identity-based verification straightforward. Packages
with stochastic output, compiled code, or side effects would not permit the
verification discipline that made this work trustworthy.

# Conclusion

An LLM agent working on a released CRAN package reproduced 14 latent
defects and produced verified-identical speed-ups of up to 20.8 times on
the package's most-used code path, while the correctness fixes themselves
imposed a measurable slowdown on two query functions that the optimisation
work subsequently recovered. It also introduced four behavioural
regressions, all caught by strict automated comparison, and made several
confident false claims about which code mattered, none of which any
automated check could have caught and all of which required the
maintainer's domain knowledge to correct.

We take the practical lesson to be that these two failure modes should be
managed differently. Behavioural risk is mechanisable and should be
mechanised aggressively — strict identity comparison over randomised
inputs is cheap and caught everything in that class here. Relevance risk is
not mechanisable, and a maintainer's scarce review attention is better
spent asking whether a proposed change is worth making than whether it is
correct.

# Reproducibility

All scripts are in the package repository under `benchmarks/`. Version
comparisons use git worktrees at the three commits described in the text,
benchmarked by `benchmarks/version-benchmark.R` with a fixed seed and
repeated measurements.
