---
title: "Agent-Assisted Defect Detection and Optimisation in an R Package: A Case Study on stagedtrees"
author:
  - name: Gherardo Varando
    affiliation: Universitat de València
    email: gherardo.varando@uv.es
abstract: >
  Large language model agents are increasingly proposed for software
  maintenance, but published evidence is dominated by synthetic benchmarks
  rather than sustained work on real, released software. We report a case
  study in which an LLM agent (Claude, Anthropic) reviewed and optimised
  **stagedtrees**, a CRAN package of roughly 6,800 lines implementing staged
  event tree models. Over a single extended session the agent identified 14
  latent defects, all reproduced before being reported, and applied a series
  of performance changes with output verified identical: the package's
  central fitting routine and its query paths are several times faster, and
  its two structure-search algorithms, whose candidate selection moved into
  compiled code, are faster by two to nearly three orders of magnitude —
  turning model sizes that did not complete in ten minutes into searches of a
  few seconds. We give equal weight to the agent's failures: it introduced a
  regression caught only by the package's existing test suite, proposed an
  "obvious" optimisation that would have silently changed a stored
  attribute's type, omitted a namespace import, made confident but incorrect
  claims about which functions mattered, and — most instructively — failed
  repeatedly at measurement rather than at code: it explained in fluent
  mechanistic terms a performance difference that a corrected experimental
  design showed did not exist; it benchmarked a non-default invocation of a
  heavily used function and so reported no change where its actual usage had
  improved; it quoted a 2300-fold speed-up from a prototype that was not
  computing what the function it was compared against computes; and it
  measured compiled code through a development loader that compiles without
  optimisation, understating its own result fourfold. We argue that the
  practical finding is a distinction by how failures are caught: mechanical
  verification reliably catches behavioural regressions, whereas judgements
  about relevance, the representativeness of a benchmark, and explanations
  offered for unverified measurements remain the reviewer's responsibility.
output:
  rticles::rjournal_article:
    keep_tex: true
bibliography: references.bib
header-includes:
  - \usepackage{longtable}
  - \usepackage{booktabs}
  - \usepackage{array}
  - \usepackage{calc}
  # the RJournal template omits the table macros pandoc emits
  - \providecommand{\real}[1]{#1}
  - \providecommand{\tightlist}{\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
options(stringsAsFactors = FALSE)

bench_dir <- Sys.getenv("BENCH_DIR", unset = "../benchmarks/results")
have_bench <- dir.exists(bench_dir) &&
  length(list.files(bench_dir, pattern = "^bench-v[0-9]+\\.csv$")) >= 2

if (have_bench) {
  files <- list.files(bench_dir, pattern = "^bench-v[0-9]+\\.csv$", full.names = TRUE)
  bench <- do.call(rbind, lapply(files, read.csv))
  agg <- aggregate(seconds ~ version + workload + size, bench, median)
  wide <- reshape(agg, idvar = c("workload", "size"), timevar = "version",
                  direction = "wide")
  names(wide) <- sub("^seconds\\.", "", names(wide))
  if (all(c("v0", "v2") %in% names(wide))) {
    wide$speedup <- wide$v0 / wide$v2
    wide <- wide[order(-wide$speedup), ]
  }
  ## Versions are interleaved across repetitions, so a paired comparison on
  ## repetition is legitimate. Report an interval rather than a point ratio:
  ## the v1-vs-v0 contrast in particular must not be read off a bare median.
  ci_ratio <- function(wl, a, b) {
    d <- bench[bench$workload == wl, ]
    w <- reshape(d[, c("version", "rep", "seconds")], idvar = "rep",
                 timevar = "version", direction = "wide")
    x <- w[[paste0("seconds.", a)]]; y <- w[[paste0("seconds.", b)]]
    ok <- stats::complete.cases(x, y)
    if (sum(ok) < 3) return(NULL)
    tt <- stats::t.test(y[ok], x[ok], paired = TRUE)
    list(ratio = median(y[ok]) / median(x[ok]), p = tt$p.value,
         lo = tt$conf.int[1], hi = tt$conf.int[2], n = sum(ok))
  }

  ## Per-cell paired contrast with a Holm correction across the whole sweep.
  ## The narrative below is computed from this rather than written out, so a
  ## re-run cannot leave the prose asserting a result the data no longer
  ## shows -- which is exactly what happened to an earlier draft.
  contrast <- function(a, b) {
    bench$key <- paste(bench$workload, bench$size)
    rows <- lapply(split(bench, bench$key), function(sdf) {
      x <- sdf[sdf$version == a, ]; y <- sdf[sdf$version == b, ]
      x <- x[order(x$rep), "seconds"]; y <- y[order(y$rep), "seconds"]
      if (length(x) != length(y) || length(x) < 3) return(NULL)
      pv <- if (all(x == y)) 1 else stats::t.test(y, x, paired = TRUE)$p.value
      data.frame(workload = sdf$workload[1], size = sdf$size[1],
                 a = median(x), b = median(y), p = pv)
    })
    out <- do.call(rbind, rows)
    out$holm <- stats::p.adjust(out$p, "holm")
    out$pct <- 100 * (out$b / out$a - 1)
    out[order(out$pct), ]
  }
  cmp02 <- contrast("v0", "v2")
  cmp01 <- contrast("v0", "v1")
  n_better <- sum(cmp02$holm < 0.05 & cmp02$pct < 0)
  n_cells <- nrow(cmp02)
  n_flat <- n_cells - sum(cmp02$holm < 0.05)
  best <- cmp02[which.min(cmp02$pct), ]
  fix_sig <- cmp01[cmp01$holm < 0.05, ]
}
```

# Introduction

Automated program repair and LLM-based coding assistants are an active
research area, but the evidence base is skewed. Benchmarks such as SWE-bench
measure an agent's ability to reproduce a known patch for a known issue;
they say little about the open-ended task a package maintainer actually
faces, which is to find defects nobody has reported yet and to make code
faster without changing what it computes.

This paper reports a case study of that open-ended task. The subject is
**stagedtrees** [@carli2022], a CRAN package implementing staged event tree
models — asymmetric graphical models for categorical data. The package is
roughly 6,800 lines of R across 65 files, has been on CRAN since 2020, and
is actively maintained.

We make three contributions. First, we characterise the defects an agent
found in released, tested code, and the discipline that made those findings
trustworthy. Second, we report performance work with verified-identical
output and quantify it across package versions. Third — and we regard this
as the most transferable result — we document the agent's failures in
detail. These divide by how they were caught rather than by what they
were: some by mechanical verification, and some only by a maintainer who
knew how the software is used. Two of them were failures of the
*measurement* rather than the code, which is a category we had not
anticipated and which the mechanical safeguards do not address.

A necessary caveat: this paper describes work performed by an LLM agent,
and the first draft of this manuscript was produced by that same agent.
Self-report is a weak form of evidence. We have tried to mitigate this by
making every quantitative claim reproducible from scripts in the package
repository, and by giving failures at least as much space as successes.
Section \ref{sec:limitations} returns to this.

# Setting and method

## The package

stagedtrees fits staged event trees: models in which the vertices of an
event tree are partitioned into *stages* that share a conditional
probability distribution. The computational core is a greedy search over
stage partitions, scored by a penalised likelihood.

## Protocol

Work proceeded in three phases over one extended session:

1. **Review.** Four parallel sub-agents examined disjoint parts of the
   codebase. The governing instruction was: *do not report a finding you
   have not reproduced.* Every candidate defect had to be demonstrated with
   a concrete failing input before it could be reported.
2. **Triage.** The maintainer approved findings individually. Each approved
   fix became a separate commit with a regression test, so that any single
   fix could be reverted without the others.
3. **Optimisation.** Profiling directed the work; every change was required
   to demonstrate identical output, not merely similar output.

The third requirement did most of the load-bearing work, and is discussed
in Section \ref{sec:failures}.

# Defects found

Fourteen defects were reproduced and fixed. They fall into recognisable
classes rather than being idiosyncratic.

## Sequence-construction errors

Seven defects shared one root cause: `1:n` where `seq_len(n)` was meant.
When `n` is zero, `1:0` yields `c(1, 0)` and the loop runs twice with an
out-of-range index, rather than not running at all. In stagedtrees this
produced a crash in `sample_from()` on a single-variable model, a crash in
`prob()` on zero-row input, and — most seriously — *silent* double
iteration in two expectation-maximisation loops, where `for (i in
1:max_iter)` with `max_iter = 0` performed two EM iterations and changed
the fitted probabilities rather than leaving them untouched.

This class is worth highlighting because it is invisible to ordinary
testing: the failure requires a boundary input (an empty vector, a
single-variable model, a zero iteration count) that a test suite organised
around typical usage will never generate.

## Silent numerical corruption

Three defects produced wrong numbers rather than errors.

`prob()` returned `1` for every conditional query on a zero-probability
conditioning event. The mechanism: the probability was computed in log
space, where `log(0) - log(0)` gives `NaN`; a guard intended to map missing
values to zero mapped that `NaN` to `0`, and `exp(0)` is `1`. A query whose
correct answer is undefined returned the maximum possible probability.

`rename_stage()` silently merged two stages when the target name already
existed, overwriting one stage's probability vector with another's and
leaving the cached log-likelihood stale.

`indep.data.frame()` dropped missing values per-variable while the
contingency-table constructor used complete cases, so the same model fitted
through two code paths reported different log-likelihoods.

## Structural assumptions

The largest single fix concerned an undocumented convention. The `$stages`
component is a named list with one entry per non-root variable, but a root
entry is *optional* — some constructors add it, others do not. Five
functions indexed `$stages` positionally, and therefore computed silently
wrong results whenever the optional entry was present. Degrees of freedom,
and hence every BIC-based model comparison, were affected.

# Performance

```{r perf-table, results = "asis"}
if (have_bench && "speedup" %in% names(wide)) {
  tab <- wide[, c("workload", "size", "v0", "v1", "v2")]
  tab$fix_cost <- sprintf("%+.1f%%", 100 * (wide$v1 / wide$v0 - 1))
  tab$speedup <- sprintf("%.2f", wide$speedup)
  names(tab) <- c("Workload", "Size", "v0 (s)", "v1 (s)", "v2 (s)",
                  "v1 vs v0", "v0/v2")
  print(knitr::kable(tab, digits = 3, row.names = FALSE, booktabs = TRUE,
    caption = paste("Median elapsed time over repeated runs.",
      "v0 is the package before this work, v1 after the defect fixes,",
      "v2 after the optimisations.")))
} else {
  cat("_Benchmark results not found; run `benchmarks/run-versions.sh` first._\n")
}
```

```{r perf-prose, results = "asis"}
if (have_bench && exists("cmp02")) {
  cat(sprintf(
"Of the %d workload/size cells, %d improve with a Holm-corrected $p < 0.05$
on a paired test over repetitions, and %d show no significant change. The
unchanged cells are the smallest ones, where a single call takes a few
milliseconds and the clock resolution dominates; they are retained to show
that nothing regressed, not as evidence of improvement. The largest effect
is `%s` at %s, from %.2f s to %.3f s.

",
    nrow(cmp02), sum(cmp02$holm < 0.05 & cmp02$pct < 0),
    nrow(cmp02) - sum(cmp02$holm < 0.05),
    best$workload, sub(" \\(.*", "", best$size), best$a, best$b))

  sig <- cmp01[cmp01$holm < 0.05, ]
  if (nrow(sig) == 0) {
    cat("The v1 column is not expected to differ from v0, since no
performance work had been done at that point, and after correcting for
multiple comparisons none of the cells does.

")
  } else {
    cat(sprintf(
"The v1 column is not expected to differ from v0, since no performance work
had been done at that point. After correcting for multiple comparisons
%d of the %d cells nonetheless %s: %s. Section \\ref{sec:artefact} describes
what happened when we drew conclusions from an uncorrected contrast of this
kind.

",
      nrow(sig), nrow(cmp01), if (nrow(sig) == 1) "does" else "do",
      paste(sprintf("`%s` at %s (%+.0f%%)", sig$workload,
                    sub(" \\(.*", "", sig$size), sig$pct), collapse = "; ")))
  }
}
```

The single large v1 effect is worth dwelling on, because its cause is not in
the code under test. `stages_hclust` checked whether **fastcluster** was
available using `rlang::is_installed()`, which costs about 0.125 s the first
time it is called in an R session and 0.001 s thereafter. An interactive user
pays that once and never notices it. The benchmark runs every measurement in
a fresh process, so it paid the cost on every repetition — and because the
cost is a constant, it swamped the four-variable model, where the entire
search takes 0.07 s, while vanishing into the noise of the six-variable one,
where the search takes 3 s. What looked like a size-dependent algorithmic
regression was a fixed session-startup cost, visible only because of how the
measurement was structured. Replacing the call with `requireNamespace()`,
which costs 0.003 s and loads the namespace that is used on the next line in
any case, removes it.

Profiling, rather than intuition, directed every optimisation. Three
findings dominated.

**Fitting grouped situations quadratically.** `sevt_fit()` assigned
situations to stages by scanning the entire stage vector once per stage.
This is $O(\text{stages} \times \text{situations})$, and a saturated model
has one stage per situation, so the deepest variable of an eight-variable,
four-level model performed $16384^2 \approx 2.7 \times 10^8$ string
comparisons. Partitioning in a single pass with `split()` removed the
quadratic term. Because `full()` is the entry point for essentially every
workflow in the package, this was the most valuable change made.

**The search evaluated candidates by materialising them.** The backward
hill-climbing search scored each candidate stage merge by constructing a
complete modified model object — copying the stage vector, rebuilding the
nested probability list, carrying the contingency tables — and then reading
two scalars out of it, discarding the object for every candidate but the
winner. Since the score is a function of the log-likelihood and the degrees
of freedom alone, and both change by an amount computable in closed form
from the two probability vectors being merged, candidates can be scored
without building anything.

This required an interface decision, and the first resolution was the
wrong one. The `score` argument is user-supplied and receives a whole
model, so a closed-form fast path cannot be applied to an arbitrary score.
The agent proposed letting `score` be *either* a string naming a predefined
score (`"BIC"` or `"AIC"`) or a function, with strings taking the fast path
and functions retaining the original behaviour. That works, but it buys
speed only for users who give up the generality of the argument, and it
adds a registry of predefined scores to the public interface.

The maintainer proposed a better decomposition: *select* the candidate by
log-likelihood alone, which is score-independent and cheap, then *accept or
reject* that single candidate with the user's score. This is valid because
every pairwise merge reduces the degrees of freedom by exactly $k-1$, where
$k$ is the number of levels of the variable. Candidates therefore differ
only in log-likelihood, and any score that is increasing in the
log-likelihood at fixed complexity — which includes AIC, BIC and every
penalised-likelihood score in common use — ranks them identically. The
score is evaluated once per accepted step rather than once per candidate,
the argument stays an ordinary function, and the registry disappears. The
documentation states the one assumption this makes, so that a score
inspecting the stage structure itself rather than the fitted log-likelihood
and dimension is known to fall outside the guarantee.

The same decomposition extends to the forward search, `stages_hc()`, with
one complication: moving a single situation between stages changes the
degrees of freedom by $-(k-1)$, $0$, or $+(k-1)$ depending on whether the
move dissolves its source stage, leaves both stages populated, or creates a
new one. Candidates are therefore not comparable by log-likelihood across
those three cases. Grouping them by the change in degrees of freedom
restores comparability within each group, and scoring the best
representative of each of the three groups — at most three score
evaluations per step instead of one per candidate — recovers the same
result.

**Index computation recomputed model-invariant quantities.** The function
mapping a root-to-leaf path to an integer index recomputed the vector of
variable arities on every call — 17 of the 65 microseconds a call took —
for a quantity fixed by the model, and rebuilt a stride vector that is a
cumulative product using a quadratic pattern.

## What Rcpp was, and was not, worth

The package originally contained no compiled code, so we measured whether
it should. A no-op compiled function called from R costs approximately 3
microseconds on the test machine. This exceeds the work done by any of the
individual helper functions identified above: a C++ implementation of the
path-index function measured 2.70 microseconds against 12 microseconds for
an optimised R implementation — a fourfold gain that would purchase a
`src/` directory, a `LinkingTo` dependency and a permanent cross-platform
build burden, and which we declined.

The calculus inverts when an entire loop moves into C rather than a single
function, because the per-call overhead is then paid once instead of once
per iteration. The actionable rule is *port loops, not leaves*. Applying it,
we ported the candidate-selection loops of the two search functions —
the only two places where the loop is both hot and closed-form — and left
every leaf function in R.

The two kernels compute, for each candidate merge or move, the change in
log-likelihood it would produce, and return only the best candidate; R
retains the model manipulation, the scoring, the acceptance decision and
the entire user-facing interface. The table below gives the result for the
installed package. The gains reach nearly three orders of magnitude, and
change what is feasible rather than merely what is fast: backward
hill-climbing over seven variables did not complete within ten minutes
before and takes about eight seconds after, and forward hill-climbing did
not complete at five variables, where it now takes one second.

```{r compiled-table, echo = FALSE, results = "asis"}
cf <- "../benchmarks/results/compiled.csv"
capf <- "../benchmarks/results/compiled-cap.txt"
if (file.exists(cf)) {
  cd <- read.csv(cf, stringsAsFactors = FALSE)
  cap <- if (file.exists(capf)) as.numeric(read.csv(capf)$cap[1]) else NA
  wide <- reshape(cd[, c("version", "fun", "p", "seconds")],
                  idvar = c("fun", "p"), timevar = "version",
                  direction = "wide")
  fmt <- function(x) ifelse(is.na(x), sprintf("> %g", cap), sprintf("%.3f", x))
  ## a capped cell gives a LOWER bound on the ratio, not a missing value
  ratio <- ifelse(is.na(wide$seconds.rver),
                  sprintf("> %.0f", cap / wide$seconds.cver),
                  sprintf("%.0f", wide$seconds.rver / wide$seconds.cver))
  tab <- data.frame(Function = wide$fun, Variables = wide$p,
                    R = fmt(wide$seconds.rver), C = fmt(wide$seconds.cver),
                    Ratio = ratio)
  tab <- tab[order(tab$Function, tab$Variables), ]
  names(tab) <- c("Function", "Variables", "pure R (s)", "compiled (s)",
                  "speed-up")
  print(knitr::kable(tab, row.names = FALSE, booktabs = TRUE,
    caption = paste("Elapsed time before and after the compiled selection",
      "kernels, 5000 observations and 3 levels per variable, with the",
      "package installed at the default optimisation level. Entries marked",
      sprintf("> %g did not complete within that many seconds.", cap))))
} else {
  cat("_Compiled-kernel results not found; run",
      "`benchmarks/run-compiled.sh` first._\n")
}
```

Two findings from this port generalise beyond the package.

**Measure the installed package, not the development load.** The agent's
first end-to-end figures for the kernels were produced under
`pkgload::load_all()`, which is how the package is loaded during
development. `load_all()` compiles with the debug flags, `-O0`; `R CMD
INSTALL` uses `-O2`. Head-to-head on identical inputs producing identical
answers, the same kernel took 11.80 ms under `load_all()` and 3.02 ms
compiled normally, a factor of 3.90. Every compiled measurement taken
during development was therefore understating the result users would see,
and the benchmark harness had to be changed to
install each version under test rather than load it. The direction of this
error is benign — it understates rather than overstates — but the magnitude
is large enough to invert a decision about whether a port is worth making.

**Rewriting in C changes the arithmetic, not just the speed.** R's `sum()`
accumulates in extended precision; C `double` does not. The first kernel
disagreed with its R oracle on which merge to select. The proximate fix was
to accumulate in `long double`, which restored agreement at roughly four
times the cost. The actual defect, however, was not accumulation error but
*catastrophic cancellation*: the delta was being formed as the difference
of two large, nearly equal log-likelihoods, so when the true difference is
zero — two stages with identical probability vectors, which balanced data
produces readily — the result is pure rounding residue, and that residue
decides the comparison. Summing the per-level differences in a single pass
removes the cancellation rather than out-precisioning it. Over 72 sweeps
against an extended-precision reference: `long double` 12.21 ms with no
disagreements, plain `double` formed as two sums 3.09 ms with one
disagreement, and the fused single sum 3.19 ms with no disagreements. The
numerically better formulation is also, here, the fast one.

The tie-breaking rule is a related trap. R's `which.max`-style loops in
this package use `>=`, so the last of several tied candidates wins; the
first C implementation used `>`, so the first won. On balanced data all 28
candidate pairs for a variable can carry an identical delta, and the two
rules then select different models — a difference that no test on random
data would have exposed, and that surfaced only because the maintainer
asked what happens under an arbitrary score.

# Where the agent was wrong {#sec:failures}

The agent's errors divide into two classes, and the distinction between
them is the practical finding of this study.

## Class 1: behavioural regressions, caught mechanically

Four errors changed program behaviour and were all caught by automated
means before reaching the repository.

**An omitted namespace import.** A fix introduced `complete.cases()` and
`setNames()` without declaring them, which `R CMD check` reported as
undefined globals. Caught by the standard package check.

**A reintroduced boundary bug.** The rewritten path-index function indexed
`path[[k]]` with `k = 0` for an empty path, where the previous
implementation returned `NA`. This is notable: the agent had verified the
rewrite exhaustively over all 396 paths of four trees, and the check missed
the case because the enumeration used `expand.grid()`, which never produces
a zero-length path. The regression was caught by a test the *same agent*
had added earlier in the session, for a defect in the same class. An
exhaustive check is only as good as its enumeration.

**A plausible optimisation that was neither.** Replacing `apply(x, 2, sum)`
with `colSums(x)` is a standard R idiom. Here it gained nothing — measured
at 1.0 times, because the cost was the surrounding scan, not the
summation — and `colSums()` returns double where `apply()` returns integer,
which would have silently changed the storage type of a sample-size
attribute on every fitted probability vector. Caught by comparing 192
fitted models with `identical()` rather than `all.equal()`.

**A dropped incidental attribute.** A vectorised replacement for a
row-filling loop lost names that the original `array()` call had
incidentally attached to a `dim` attribute. Semantically irrelevant;
caught by the same strict comparison.

The common thread is that strict structural comparison over many inputs
caught all four, whereas value comparison (`all.equal`) would have caught
only the first two. The cost of that strictness was low and its yield was
high.

## Class 1b: a fabricated result, from the measurement itself {#sec:artefact}

One error belongs in this class only partly, and is the one we find most
instructive, because the flawed artefact was the *experiment* rather than
the code.

An earlier version of the benchmark harness in this paper ran every
repetition of one version before proceeding to the next. Machine
performance is not stationary over a run of tens of minutes — thermal
behaviour, competing load on shared infrastructure, page-cache state all
drift — and under a version-blocked design that drift is confounded with
the version under test.

The result was an apparent 13.7% slowdown in `predict()` and 7.5% in
`prob()` between v0 and v1, with tight, non-overlapping ranges across five
repetitions. The agent accepted this, and wrote a paragraph of this paper
explaining it: the zero-probability guard had added a detection pass, and
the change from positional to name-based indexing had replaced integer
subscripting with string lookup. Both mechanisms are real features of the
diff. Neither explanation was checked against the diff before being
written, and the second is simply false — the relevant function contains no
name-based indexing at all.

The error surfaced only because an unrelated question — whether name-based
indexing was costly enough to warrant redesign — prompted a direct
comparison.

```{r artefact-table, results = "asis"}
if (have_bench) {
  ks <- unique(paste(bench$workload, bench$size))
  rows <- lapply(ks, function(k) {
    d <- bench[paste(bench$workload, bench$size) == k, ]
    w <- reshape(d[, c("version", "rep", "seconds")], idvar = "rep",
                 timevar = "version", direction = "wide")
    ## a workload missing from a version means the result set is incomplete
    ## or mixes runs; refuse to tabulate rather than report a partial row
    if (!all(c("seconds.v0", "seconds.v1") %in% names(w))) return(NULL)
    if (sum(stats::complete.cases(w$seconds.v0, w$seconds.v1)) < 3) return(NULL)
    tt <- stats::t.test(w$seconds.v1, w$seconds.v0, paired = TRUE)
    data.frame(Workload = k,
               `Change` = sprintf("%+.1f%%",
                 100 * (median(w$seconds.v1) / median(w$seconds.v0) - 1)),
               `Absolute` = sprintf("%+.4f s",
                 median(w$seconds.v1) - median(w$seconds.v0)),
               p = tt$p.value, check.names = FALSE)
  })
  tab <- do.call(rbind, rows[!vapply(rows, is.null, TRUE)])
  tab$`p (Holm)` <- sprintf("%.3f", stats::p.adjust(tab$p, method = "holm"))
  tab$p <- sprintf("%.4f", tab$p)
  print(knitr::kable(tab, row.names = FALSE, booktabs = TRUE,
    caption = paste("v1 against v0, paired by repetition. No performance",
      "work separates these versions, so differences are not expected.",
      "Holm correction over the nine tests.")))
}
```

Re-measured with the versions interleaved across repetitions, the two
functions named in the retracted paragraph are firmly null: `predict()`
differs by $+0.8\%$ ($p = 0.69$) and `prob()` by $+0.6\%$ ($p = 0.24$).
Profiles of the two versions are structurally identical. The 13.7% was an
artefact, and the mechanism invented for it described functions that were
never affected.

One workload does differ. `full()` on the smallest input is 30.4% slower in
v1, and this survives correction for the nine comparisons
($p_{\text{Holm}} = 0.003$). The absolute difference is 19 milliseconds on
a 62 millisecond call, which is consistent with the correctness fixes
having added fixed per-call cost that is only visible when the call is
short — but we have not established that, and say so rather than supplying
a second mechanism on the strength of a second unexplained number. It is
noted here as an open observation.

That the surviving effect is on a workload the original claim never
mentioned, while both workloads it did name are null, is the point. The
agent's account was not an exaggeration of a real signal; it was
independent of the data.

## Class 1c: measuring the wrong call

A third measurement failure, found the same way — by the maintainer
remarking in passing which algorithms users actually run.

The benchmark exercised `stages_hclust(m, k = 3)`, fixing the number of
stages. The function's default is `k = NA`, under which it searches over
the number of stages, refitting the model once per candidate. That default
path costs roughly thirty times as much as the fixed-`k` call, and it is
the path that benefits from the fitting improvements. Both are now measured
as part of the benchmark, and the contrast is stark:

```{r hclust-table, results = "asis"}
if (have_bench) {
  hk <- grep("hclust", unique(paste(bench$workload, bench$size)), value = TRUE)
  rows <- lapply(hk, function(k) {
    d <- bench[paste(bench$workload, bench$size) == k, ]
    w <- reshape(d[, c("version", "rep", "seconds")], idvar = "rep",
                 timevar = "version", direction = "wide")
    tt <- stats::t.test(w$seconds.v2, w$seconds.v0, paired = TRUE)
    data.frame(Call = sub(" 2000x6x3", "", k),
               v0 = sprintf("%.3f s", median(w$seconds.v0)),
               v2 = sprintf("%.3f s", median(w$seconds.v2)),
               Change = sprintf("%+.1f%%",
                 100 * (median(w$seconds.v2) / median(w$seconds.v0) - 1)),
               p = sprintf("%.4f", tt$p.value), check.names = FALSE)
  })
  print(knitr::kable(do.call(rbind, rows), row.names = FALSE, booktabs = TRUE,
    caption = paste("stages\\_hclust under its default and with k fixed.",
      "Only the default was affected by the optimisation work, and only",
      "the default is what users call.")))
}
```

The benchmark had therefore reported "no change" for one of the two
algorithms the maintainer identifies as most used, while its default
invocation had improved by 27%. The error is not statistical — the
fixed-`k` measurement is perfectly sound — but one of construction: the
workload did not represent how the function is called. A fixed `k` was
chosen, so far as can be reconstructed, because it made the workload
faster and more deterministic, which are properties of a convenient
benchmark rather than a representative one.

This differs from the artefact of Section \ref{sec:artefact} in an
important way. There, the measurement was invalid and produced a false
signal. Here every number was correct; the workload simply answered a
question nobody had asked. False signals are at least detectable by
re-measurement. A valid measurement of an unrepresentative case is not,
because there is nothing internally wrong with it to find — it is caught
only by someone who knows how the software is used, which returns us to
the division of labour argued in Section \ref{sec:discussion}.

We draw two lessons. The first is mundane and well known outside software
benchmarking: block on the nuisance variable, or randomise against it.
The harness now interleaves versions, and carries a comment recording why.

The second is specific to agent-assisted work. The agent's failure was not
the flawed design — that is an ordinary methodological slip — but the
readiness to explain the artefact. Presented with a number, it produced a
fluent, plausible, partly false mechanism for it within seconds, and that
explanation is materially more persuasive than the evidence warranted. An
agent that generates hypotheses cheaply will generate them for noise as
readily as for signal, and the resulting prose carries no marker
distinguishing the two. The discipline that protects code — verify the
claim against the artefact before asserting it — applies verbatim to
measurement, and was not applied here until prompted.

## Class 2: misjudged relevance, caught only by the maintainer

The second class is more troubling, because no test could have caught it.

The agent profiled `stages_bj()` at 21.6 seconds and identified a genuine
superquadratic scaling defect in it — a distance matrix rebuilt in full
after every merge when only one row changes. It recommended the fix. It
also asserted that related changes would accelerate `stages_hclust()`.

The maintainer's response was that `stages_hclust()` uses R's built-in
`dist()` for its documented distance options and never calls the affected
code at all, and that the paths in question are not ones users exercise.
Both points were correct; the agent's claim about `stages_hclust()` was
simply false, and verifiable in seconds by grep — which the agent performed
only after being challenged, confirming the maintainer. The recommended
work was abandoned.

The agent had made a similar error earlier, reporting that four search
functions shared a code pattern when one of them — which takes no score
argument at all — did not.

These were not failures of reasoning about code behaviour, which the agent
did reliably. They were failures of judgement about *what matters*: which
functions are used, which costs users actually pay, where optimisation
effort earns its risk. That knowledge lives in the maintainer's experience
of the user base and is not recoverable from the source tree.

The redirection was consequential. Abandoning `stages_bj()` and
investigating `full()` instead led directly to the largest result in this
paper.

# Discussion {#sec:discussion}

The two failure classes suggest an asymmetric division of labour.

For behavioural correctness, the agent was effective and mechanical
verification was sufficient. The productive discipline was not
sophisticated: reproduce before reporting, compare with `identical()`
rather than `all.equal()`, compare across many randomised inputs rather
than one, and treat any diagnosed difference as a reason to stop rather
than a detail to explain away. Every Class 1 error was caught this way,
and the cost was a few minutes of compute per change.

For relevance, the agent was actively misleading, because its errors were
delivered with the same confidence as its correct findings. An agent that
profiles a function at 21.6 seconds and proposes a correct algorithmic fix
gives no signal that the function is rarely called. Nothing in the
repository encodes that. A maintainer reading the same report identifies it
immediately.

This suggests that the review burden on a human collaborator does not scale
down in proportion to the agent's competence. The agent can be trusted to
establish *that* a change is safe; it cannot be trusted to establish that
the change is *worth making*, and the second question is where the
maintainer's attention is best spent.

# Limitations {#sec:limitations}

This is a single case study, and its limitations are substantial.

**Self-report.** The work was performed by an LLM agent and the first draft
of this paper was written by that agent. An agent summarising its own
performance is a poor instrument, particularly for detecting errors it did
not notice making. The failures documented in Section \ref{sec:failures}
are those that were *caught*; failures that passed both automated checks
and maintainer review would by construction be absent from this account.

**No control.** We do not know what a human reviewer, a static analyser, or
a different agent would have found in the same code. The defect count is
therefore not a measure of anything comparative. We note that several
defects — particularly the `1:n` family — are exactly what existing lint
tooling targets, and would plausibly have been found by cheaper means.

**Benchmark conditions.** Timings were collected in a single containerised
environment on one machine. Absolute figures should not be compared across
systems; the ratios are the meaningful quantity, and even those carry
run-to-run variance that the repeated-measures design exposes but does not
eliminate.

**Defect severity is unmeasured.** The defects found were latent: none
corresponded to a user-reported failure. Their practical importance is
therefore unknown, and counting them conflates a silent EM-iteration bug
with a redundant loop.

**Generalisation.** One package, one domain, one agent, one session. The
codebase is mid-sized and has an unusually well-defined numerical contract,
which makes identity-based verification straightforward. Packages with
stochastic output or substantial side effects would not permit the same
discipline. Compiled code is a partial exception rather than a flat
obstacle: the kernels added here were verified against a pure-R oracle
retained in the package for that purpose, which worked, but it required
writing the oracle first and it exposed a class of divergence — floating
point differences in the last bits, deciding a comparison between near-tied
candidates — that has no analogue in R-to-R refactoring. Identity is the
right standard within one arithmetic; across two it has to be argued for.

# Conclusion

An LLM agent working on a released CRAN package reproduced 14 latent
defects and produced verified-identical speed-ups across the package's
most-used code paths, reaching two to nearly three orders of magnitude on
the structure-search algorithms once their candidate selection moved into
compiled code. It also introduced four behavioural regressions, all caught
by strict automated comparison, and made several confident false claims
about which code mattered, none of which any automated check could have
caught and all of which required the maintainer's domain knowledge to
correct.

The most persistent failures were not in the code but in the measurements
taken of it. Four separate times a number was produced that was
wrong not because the arithmetic was wrong but because the experiment
answered a different question than the one being asked: a version-blocked
design that confounded machine drift with the version under test; a
benchmark of a non-default invocation; a prototype that omitted the work it
was being credited with doing; and a development-time loader that compiles
without optimisation. Each was caught by re-measuring rather than by
reasoning, and one of them — a fixed session-startup cost that the
harness paid once per repetition — masqueraded convincingly as a
size-dependent algorithmic regression.

We take the practical lesson to be that these two failure modes should be
managed differently. Behavioural risk is mechanisable and should be
mechanised aggressively — strict identity comparison over randomised
inputs is cheap and caught everything in that class here. Relevance risk is
not mechanisable, and a maintainer's scarce review attention is better
spent asking whether a proposed change is worth making than whether it is
correct.

# Reproducibility

All scripts are in the package repository under `benchmarks/`. Version
comparisons use git worktrees at the three commits described in the text,
benchmarked by `benchmarks/version-benchmark.R` with a fixed seed and
repeated measurements.
